Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-15 talk/iwtc11/paper.tex
remove is_positive fully

For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
 a very simple object
model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
two operations, \lstinline{add}, which adds two objects (promoting ints to floats in a
mixed addition) and \lstinline{is_positive}, which returns whether the number is greater
than zero. The implementation of \lstinline{add} uses classical Smalltalk-like
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

@@

   def is_positive(self):
      return self.intval > 0

@@

   def is_positive(self):
      return self.floatval > 0.0

def f(y):
   step = BoxedInteger(-1)
   while y.is_positive():
      y = y.add(step)
   return res
\end{lstlisting}
\caption{An ``Interpreter'' for a Tiny Dynamic Language Written in RPython}
\label{fig:objmodel}
\end{figure}

@@

Let us now consider a simple ``interpreter'' function \lstinline{f} that uses the
object model (see the bottom of Figure~\ref{fig:objmodel}).
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, one for each \lstinline{is_positive} and even two for each
call to \lstinline{add}. These method calls need to check the type of the involved
objects every iteration. In addition, a lot of objects are created
when executing that loop, many of these objects are short-lived.
The actual computation that is performed by \lstinline{f} is simply a sequence of
float or integer additions.

\begin{figure}
XXX the code for is\_positive is missing everywhere
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2} + i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_0$, $p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}

>>>>

For the purpose of this paper, we are going to use a tiny interpreter for a dynamic language with
 a very simple object
model, that just supports an integer and a float type (this example has been taken from a previous paper \cite{bolz_allocation_2011}). The objects support only
two operations, \lstinline{add}, which adds two objects (promoting ints to floats in a
mixed addition). The implementation of \lstinline{add} uses classical Smalltalk-like
double-dispatching.
%These classes could be part of the implementation of a very
%simple interpreter written in RPython.
The classes can be seen in
Figure~\ref{fig:objmodel} (written in RPython).

@@

@@

def f(y):
   step = BoxedInteger(-1)
   while True:
      y = y.add(step)
\end{lstlisting}
\caption{An ``Interpreter'' for a Tiny Dynamic Language Written in RPython}
\label{fig:objmodel}
\end{figure}

@@

Let us now consider a simple ``interpreter'' function \lstinline{f} that uses the
object model (see the bottom of Figure~\ref{fig:objmodel}).
Simply running this function is slow, because there are lots of virtual method
calls inside the loop, two for each
call to \lstinline{add}. These method calls need to check the type of the involved
objects every iteration. In addition, a lot of objects are created
when executing that loop, many of these objects are short-lived.
The actual computation that is performed by \lstinline{f} is simply a sequence of
float or integer additions (note that \lstinline{f} does not actually terminate,
but it is still instructive to look at the produced traces).

\begin{figure}
\begin{lstlisting}[mathescape,numbers = right,basicstyle=\setstretch{1.05}\ttfamily\scriptsize]
$l_0$($p_{0}$, $p_{1}$):
# inside f: y = y.add(step)
guard_class($p_{1}$, BoxedInteger)
    # inside BoxedInteger.add
    $i_{2}$ = get($p_{1}$, intval)
    guard_class($p_{0}$, BoxedInteger)
        # inside BoxedInteger.add__int
        $i_{3}$ = get($p_{0}$, intval)
        $i_{4}$ = $i_{2} + i_{3}$
        $p_{5}$ = new(BoxedInteger)
            # inside BoxedInteger.__init__
            set($p_{5}$, intval, $i_{4}$)
jump($l_0$, $p_{0}$, $p_{5}$)
\end{lstlisting}
\caption{An Unoptimized Trace of the Example Interpreter}
\label{fig:unopt-trace}
\end{figure}
